\begin{abstract}
Plant diseases pose a significant threat to global food security, necessitating automated detection systems that are both accurate and deployable in real-world agricultural settings. This paper presents a comprehensive comparative evaluation of twelve pretrained deep learning architectures for binary plant disease classification (healthy vs.\ diseased), including classical models (AlexNet, VGG16), lightweight architectures (MobileNet\_v2, ShuffleNet\_v2, SqueezeNet, MnasNet, EfficientNet-Lite4), and modern designs (ResNet50, DenseNet121, Xception, InceptionV4, ConvNeXt). Using transfer learning on the PlantVillage dataset with targeted data augmentation to address class imbalance, we evaluate model performance through accuracy, precision, recall, F1-score, and AUC-ROC metrics. Critically, we assess generalization capability through cross-dataset evaluation on PlantDoc, which contains real-world images with natural backgrounds. Our results reveal that architectural complexity does not guarantee superior performance: VGG16 achieved the highest AUC-ROC (0.9990) on PlantVillage, while EfficientNet-Lite4 attained the best accuracy (98.20\%). However, cross-dataset evaluation exposed significant generalization gaps, with ConvNeXt demonstrating the best transfer performance (83.90\% on PlantDoc) despite ranking lower on PlantVillage. Lightweight models exhibited severe overfitting, with MobileNet\_v2 and ShuffleNet\_v2 dropping to 44.49\% and 40.25\% accuracy respectively on PlantDoc. For practical deployment, AlexNet emerges as a recommended choice, offering balanced performance (96.53\% PlantVillage, 81.36\% PlantDoc) with the fastest inference speed (651 samples/second). Our findings underscore the importance of cross-dataset evaluation for robust plant disease detection systems.
\end{abstract}